\textbf{Wavelet Scattering Tranforms Applied to Whale Vocalisations}

### 2024-05-14 dissertation/content/nomenclature.tex
started intro to p3

\begin{tabular}{p{0.15\textwidth} p{0.75\textwidth}}
    $X[k,m]$ & short-time Fourier transform of $x$, with $k$ the frequency index and $m$ the time index \\
    $S[k,m]$ & power spectrum calculated from $X[k, m]$ \\
    $H[m]$ & spectral entropy calculated from $S[k, m]$ \\
    $X(s, b)$ & continuous wavelet transform of $x$, with $s$ as scale and $b$ as translation \\
    $\uppsi_\lambda$ & sometimes used as short-hand for  $\uppsi(\lambda t)$\\
    $\text{RM}^F_m (x)$ & rolling median filter operator with a window of size $F$ operating on index variable $m$ on a discrete function $x$\\
    $\text{RA}^F_m (x)$ & rolling average filter operator with a window of size $F$ operating on index variable $m$ on a discrete function $x$\\
\end{tabular}

@@

\section*{Acronyms}
\begin{tabular}{p{0.15\textwidth} p{0.75\textwidth}}
    GMM & Gaussian mixture model \\
    POI & point(s) of interest \\
    EM & expectation maximization \\
    SE & spectral entropy \\
    PAM & passive acoustic monitoring \\
    PSD & power spectral density \\
    STFT & short-time Fourier transform \\
    TPR & true positive rate \\
    FPR & false positive rate \\
    PDF & probability density function \\
    SNR & signal-to-noise ratio \\
    AMFM & amplitude and frequency modulated \\
    BLED & band-limited energy detection \\
    TF & time-frequency \\
    CWT & continuous wavelet transform \\
    DWT & discrete wavelet transform \\
    SWT & stationary wavelet transform \\
    WPD & wavelet packet decomposition \\
    MODWT & maximal overlap discrete wavelet transform \\
    MFCC & Mel-frequency cepstral coefficient \\
    DNN & deep neural network \\
    CNN & convolutional neural network \\
    HMM & hidden Markov model \\
    SOTA & state-of-the-art \\
    LSTM & long-short-term memory \\
    EWT & Endangered Wildlife Trust \\
    WS & wavelet scattering \\
    WST & wavelet scattering transform \\
    FFT & fast Fourier transform \\
    GPU & graphics processing unit \\
    ROC & receiver operating characteristic \\
    CLS & cosine-log scattering \\
    LDA & linear discriminant analysis \\
    ML & machine learning \\
    NN & neural network \\
    SVM & support vector machine \\
\end{tabular}

@@

    1D & 1-dimensional \\
    2D & 2-dimensional \\
    SIFT & scale-invariant feature transform \\
    LPF & low-pass filter \\
    IR & impulse response \\
    FIR & finite impulse response \\
    AUC & area under curve \\
    SD & standard deviation \\
    STPR & soft true positive ratio \\
    STNR & soft true negative ratio \\

>>>>

\begin{tabular}{p{0.15\textwidth} p{0.75\textwidth}}
    $X[k,m]$ & a time-frequency decomposition of $x$, with $k$ the frequency index and $m$ the time index \\
    $S[k,m]$ & power spectrum calculated from $X[k, m]$ \\
    $H[m]$ & spectral entropy calculated from $S[k, m]$ \\
    $X(s, b)$ & continuous wavelet transform of $x$, with $s$ as scale and $b$ as translation \\
    $\uppsi_\lambda$ & sometimes used as short-hand for  $\uppsi(\lambda t)$\\
    $\text{RM}^F_m (x)$ & rolling median filter operator with a window of size $F$ operating on index variable $m$ on a discrete function $x$\\
    $\text{RA}^F_m (x)$ & rolling average filter operator with a window of size $F$ operating on index variable $m$ on a discrete function $x$\\
    $x \triangleq y$ & $x$ is defined as $y$\\
\end{tabular}

@@

\section*{Acronyms}
\begin{tabular}{p{0.15\textwidth} p{0.75\textwidth}}
    GMM & Gaussian mixture model \\
    DT & discrete-time \\
    POI & point(s) of interest \\
    EM & expectation maximization \\
    SE & spectral entropy \\
    PAM & passive acoustic monitoring \\
    PSD & power spectral density \\
    STFT & short-time Fourier transform \\
    TPR & true positive rate \\
    FPR & false positive rate \\
    PDF & probability density function \\
    SNR & signal-to-noise ratio \\
    AMFM & amplitude and frequency modulated \\
    BLED & band-limited energy detection \\
    TF & time-frequency \\
    CWT & continuous wavelet transform \\
    DWT & discrete wavelet transform \\
    SWT & stationary wavelet transform \\
    WPD & wavelet packet decomposition \\
    MODWT & maximal overlap discrete wavelet transform \\
    MFCC & Mel-frequency cepstral coefficient \\
    DNN & deep neural network \\
    CNN & convolutional neural network \\
    HMM & hidden Markov model \\
    SOTA & state-of-the-art \\
    LSTM & long-short-term memory \\
    EWT & Endangered Wildlife Trust \\
    WS & wavelet scattering \\
    WST & wavelet scattering transform \\
    FFT & fast Fourier transform \\
    GPU & graphics processing unit \\
    ROC & receiver operating characteristic \\
    CLS & cosine-log scattering \\
    LDA & linear discriminant analysis \\
    ML & machine learning \\
    NN & neural network \\
    SVM & support vector machine \\
\end{tabular}

@@

    1D & 1-dimensional \\
    2D & 2-dimensional \\
    SIFT & scale-invariant feature transform \\
    LPF & low-pass filter \\
    IR & impulse response \\
    FIR & finite impulse response \\
    AUC & area under curve \\
    SD & standard deviation \\
    STPR & soft true positive ratio \\
    STNR & soft true negative ratio \\
    BPF & band-pass filter \\

### 2024-05-14 dissertation/content/p1_intro.tex
before reference consolodation
started intro to p3

\chapter{The Relevance of Time-Frequency Analyses}

>>>>

\acrodef{dt}[DT]{discrete-time}
\acrodef{ct}[CT]{continuous-time}
\acrodef{bpf}[BPF]{band-pass filter}
\acrodef{dtft}[DTFT]{discrete-time Fourier transform}

\chapter{Time-Frequency Analyses}
\label{chap:p1i}

A \ac{tf} analysis is one of the most important tools in audio classification, data preparation and data manipulation. The main idea behind \ac{tf} decompositions is to split a signal into frequency bands, which can viewed as time-varying functions.

Formally, given a \ac{1d} \ac{dt} signal $x[n]$, with an index variable $n$, a \ac{tf} analysis decomposes $x$ as 
\begin{equation}
    X[k, n] = \sum_i x[n] \upsilon_k[i - n] = x * \upsilon_k,
\end{equation}
where $\upsilon_k \in \Upsilon$ is the $k$'th \ac{bpf} in the decomposition filter dictionary $\Upsilon$. Note that a \ac{lpf} can be considered to be a \ac{bpf} with a centre-frequency of 0.

Generally, the entire convolution $X[n,k]$ is not computed. Instead, only select samples are computed by speficying a hop-distance or downsample factor $d$. We indicate this by tranforming the index variable $m = dn$ as 
\begin{equation}
    \label{eqn:gentf}
    X[k, m] \triangleq X[k, dn] = \left(x * \upsilon_k\right)_{\downarrow d}.
\end{equation}

The is equivalent to applying a downsample operator $\left(\cdot\right)_{\downarrow d}$ to the convolution.

\section{Fourier Transforms}

A \ac{ct} Fourier transform of $x(t)$ is given as 
\begin{equation}
    \mathcal{F} \{x\} \triangleq \hat{x}(\omega) = \int_{-\infty}^{\infty} x(t) e^{-j\omega t} dt = \left(x * e^{j\omega t}\right)\bigg|_{t=0}.
\end{equation}

Derived from the conventional Fourier transform, a \ac{dt} signal $x[n], n \in \{0, ..., N-1\}$ (length $N$) has a \ac{dtft}
\begin{equation}
    \label{eqn:dtft}
    \text{DTFT}\{x\} \triangleq \hat{x}[k] = \sum_{n=0}^{N} x(t) e^{-j \frac{2\pi k n}{N}} = \left(x \otimes \upsilon_k\right)\bigg|_{n=0},
\end{equation}
for which $\upsilon_k = e^{j \frac{2\pi k n}{N}}, \ k \in \{0, ..., N-1\}$ forms an orthogonal basis of $x$. The circular convolution operator $\otimes$ implies the infinite periodic extension of $x$ beyond its boundaries of $n \in \{0, ..., N-1\}$.

Equation (\ref{eqn:dtft}) describes the \ac{dtft} as convolution evaluated at the index 0 as a useful link to the generalised notion of \ac{tf} analyses: Fourier transforms can be considered as complex sinusoidal filters being applied to the signal of interest, with periodic boundary extension.

The \ac{dtft} is commonly used to quickly compute convolutions for which the \ac{fft} is a fast alorithm with $N \log N$ computational complexity. In particular, to circularly convolve signals $x$ and $y$, a fast computation is implemented as
\begin{equation}
    \label{eqn:fftconv}
    x \otimes y = \text{FFT}^{-1}\{\hat{x} \cdot \hat{y}\}.
\end{equation}

With the correct padding procedures, one can compute $x * y$ using (\ref{eqn:fftconv}).

Fourier transforms as decribed above are invertible operators, since they are orthogonal decompositions. However, for \ac{tf} analyses, invertibility is not necessarily a requirement of the application. As such, this disseration does not consider invertibility for the remainder of the content.

\section{Constructing Filters}
\label{sec:p1i:tfa}

The most popular \ac{tf} analysis is the \ac{stft}.

In terms of equation (\ref{eqn:gentf}), a \ac{stft} is computed with 
\begin{equation}
    \label{eqn:stftfilter}
    \upsilon_k[n] =  w[n] e^{j \frac{2\pi k n}{N}},
\end{equation}
where $w$ is a window function defined by the user. Various \acp{fft} are use to compute this transform, so the properties of $w$ is typically specified by the \ac{fft} length, and the window type (Hann, Hamming, Blackman, etc.). The hop-length (or downsample factor) $d$ is an additional parameter specified by the user.

The frequency domain content $\hat{w}$ entirely specifies the filter behaviour of $\upsilon_k$, since the factor $e^{j \frac{2\pi k n}{N}}$ only serves to specify the centre frequency of $\upsilon_k$. The resulting filter dictionary $\Upsilon$ is therefore comprised of constant-bandwidth analytic (complex) filters, linearly spaced in the frequency domain. As such, depending on the definition of filter bandwidth, there exists an optimal $d$ which captures all of the necessary information of $|X[k, m]|$ such that no aliasing occurs across the index dimension $m$. This type of optimality is discussed, at length, for wavelet filterbanks in chapter \ref{chap:p3}.

The linearly-spaced filters and constant-bandwidth does not necessarily reflect and/or capture the natural phenomena typically present in audio signals. In particular, an audio signal can be expressed as a series of harmonics above its fundamental. This motivates the use of exponentially placed filters.

Formally, given a signal with a fundamental frequency most accurately captured in the frequency bin $k_0$, any change in frequency by a factor $a$ will move the fundamental frequency content to the bin $a k_0$. This has dire consequences for lower frequency signals - the frequency resolution of lower frequency signals is significantly poorer compared to the frequency resolution of its own harmonics. As such, a small frequency shift of a lower frequency signal with ample harmonic content which cause large index shifts for its harmonics, i.e., the $b$'th harmonic will be shifted to index $a b k_0$.

Suppose that we design $\Upsilon$ to contain $Q$ filters per octave. A frequency change by a factor $a$ shifts the frequency index of the fundamental to $(Q\log_2 (a) + 1) k_0$. Similarly, the $b$'th harmonic's frequency index is shifted to $(Q\log_2 (a) + 1) (Q\log_2 (b) + 1) k_0$. For example, if $a=2$, the fundamental's frequency index is shifted to $k_0 + Q$ (as opposed to $2 k_0$ for the \ac{stft}). This property is referred to a frequency-shift invariance, since small frequency deformations of the order $Q$ do not significantly change the frequency index structure of the \ac{tf} decomposition. The \ac{stft} is much more sensitive to such deformations across its entire structure for harmonically rich signals.

An exponential filterbank design has increased stability to small frequency fluctuations, which can make it more useful for feature-extraction methods in a \ac{ml} pipeline, since feature vectors are more stable against frequency deformations. The exponential filterbank is, in fact, the way wavelet filters are typically constructed. Additionally, the \ac{mfcc} filterbank attributes much of its sucess (compared to raw \ac{stft} magnitude coefficients) due to its exponentially placed filters.

\section{Wavelet Filters}
\label{sec:p1i:wavelets}

To construct a wavelet filterbank with $\upsilon_k \in \Upsilon$ which has the exponentially-spaced frequency properties discussed in section \ref{sec:p1i:tfa}, we use a ``template'' wavelet $\uppsi(t)$, referred to as the mother wavelet. Conventionally, wavelet filters are defined as \ac{ct} filters and then discretised in software implementations.

The mother wavelet is a \ac{bpf} with the following properties:
\begin{enumerate}
    \item Centre frequency of 1 rad/s.
    \item Zero mean: $\hat{\uppsi}(0) = 0$.
    \item Bandwidth defined by the user.
\end{enumerate}

$uppsi$ can the be scaled by a factor $\lambda$ such that
\begin{equation}
    \upsilon_k(t) = \lambda_k \uppsi(\lambda_k t).
\end{equation}

Notational shorthand is often used: 
\begin{equation}
    \uppsi_{\lambda_k}(t) \triangleq \lambda_k \uppsi(\lambda_k t).
\end{equation}

We construct a set $\lambda_k \in \Lambda$ such that $\lambda_{k+1} = 2^\frac{1}{Q}\lambda_{k}$, which ensures $Q$ wavelets per octave. Since $\uppsi$ has a frequency of 1 rad/s, $\lambda_k$ becomes the centre frequency of $\uppsi_{\lambda_k}$, while also scaling its bandwidth by a factor of $\lambda_k$.

An additional advantage resulting from bandwidth scaling is that higher frequencies become more localised in time, allowing for better localisation precision as frequency increases. This results from the fact that $\uppsi_{\lambda_k}$ maintains the same number of oscillations over scale, which is not true for the \ac{stft}.

Chapter \ref{chap:p1} discusses the applications of such a filterbank in terms of the \ac{cwt}, obtaining a scalogram. Note that traditional \ac{cwt} notation typically uses the scale variable $s = \frac{1}{\lambda}$, where $s$ is made continuous. Discretisations of the \ac{cwt} approximate the transform by limiting $s$ to the set $s_k \in \frac{1}{\lambda_k}, \ \lambda_k \in \Lambda$. Chapter \ref{chap:p1} specifically considers the application of the \ac{cwt} for signal detectors, while also using other techniques to improve traditional methods.

### 2024-05-14 dissertation/content/p2_intro.tex
started intro to p3

\chapter{Feature Extraction with Wavelet Transforms}

>>>>

\acrodef{dct}[DCT]{discrete cosine transform}

\chapter{Feature Extraction with Wavelet Transforms}
\label{chap:p2i}

\ac{tf} decompositions are often used to extract features for signals. In audio, \acp{mfcc} are features calculated from the \ac{stft} of the signal. A ``\ac{tf}'' decomposition, as referred to in this dissertation, does not necesarrily require a time variable: spatial variables are often used. In this sense, a \ac{cnn} is can also be considered as a ``\ac{tf}'' decomposition, where multi-dimensional index variables (or index vectors) are used for ``time'' and frequency:
\begin{equation}
    X[\vect{k}, \vect{m}] = \left( x * \upsilon_{\vect{k}} \right)_{\downarrow \vect{d}}, \ \upsilon_{\vect{k}} \in \Upsilon.
\end{equation}

In this case, we abuse notation to indicate that the operator $(\cdot)_{\downarrow \vect{d}}$ downsamples by different amounts for each dimension of $\vect{m}$.

The downsampling factor $\vect{d}$ limits the bandwidth of $X$, which, as discussed in chapter \ref{chap:p1i}, is selected so as to retain all information carried by the filters in $\Upsilon$. Although not notated, different filters (different $\vect{k}$) may have different values of $\vect{d}$. Notation to support such a case is neglected for clarity, but is later expanded upon in chapter \ref{chap:p3}.

\section{The Importance of Invariance}

The decomposition $X$ is often further processed by additional filters $\Phi_i[\vect{k}, \vect{m}]$, where $i$ is the index of the post-processing filter:
\begin{equation}
    \tilde{X_i}[\vect{k}, \vect{m}] = X * \Phi_i.
\end{equation}
The filters $\Phi_i$ are not necesarrily multi-dimensional, and may only operate along a single dimension of $X$. If $\Phi_i$ is an averaging (low-pass) filter, some additional stability can be observed in $\tilde{X_i}$ compared with $X$. This allows for the introduction of invariances to various deformations applied to $X$, which can be greatly beneficial for some applications. In a sense, cascased structures of $X$ and $\tilde{X_i}$ are defined in a \ac{cnn} for which the network learns the required invariances.

It is widely known that enforcing invariance in feature extraction and \ac{ml} pipelines increase performance. Enforcing invariance in \acp{nn} has been shown to improve the network, while also improving the interpretability of the network. For example, when classifying images, invariance to scale, shear, rotation and translation deformations, depending on image properties, may be useful when extracting features. Specifically, deformations of a specified order should approximately map to the same point in feature space:
\begin{equation}
    \Gamma\left(\delta\left(x\left[\vect{n}\right]\right)\right) \approx \Gamma\left(x\left[\vect{n}\right]\right),
\end{equation}
where $\Gamma$ is the feature extraction operator, and $\delta$ is the deformation operator applied to $x$.

Invariance filters ($\Phi_i$) can also have additional properties that makes it favourable for audio applications. For example, many applications ``blur'' (low-pass) the \ac{stft} magnitude or power spectrum, which can reduce the impact of noise fluctuations, and can also serve to suppress small frequency fluctuations across adjascent frequency bins (if the blurring kernel $\Phi_i$ filters across frequency as well). The blurring operation can improve the performance of signal detectors and classifiers. However, as a cost, time localisation and/or frequency localisation is reduced.

A new problem is introduced when enforcing invariances via the $\Phi_i$ filters - information is lost via averaging. Various techniques can be used to further retain information, while also retaining invariance. Wavelet scattering is such a method. In a sense, multilayer \acp{cnn} topologies with skip connections can be though of as a method to recover lost information in the deeper layers.

\section{MFCCs Reframed as Shift-Invariant Features}

Since \acp{mfcc} are the most widely used feature extraction method for audio signals, save for \ac{cnn} frontends which also act as a feature extractor when operating on the \ac{stft}. For this reframing, we ignore the \ac{dct} applied to the filtered values, since this linear transformation only serves to decorrelate the filter coefficients.

The magnitude spectrum used in \ac{mfcc} calculation is expressed as a \ac{tf} decomposition as defined by equations (\ref*{eqn:gentf}) and (\ref*{eqn:stftfilter}):
\begin{equation}
    X[l, m] = |\upsilon_l * x|_{\downarrow d},
\end{equation}
where $\upsilon_l$ is the $l$'th \ac{stft} filter.

The Mel spectrogram \ac{tf} decomposition $X_\text{Mel}$ is then calculated as
\begin{equation}
    X_\text{Mel}[k, m] = \log\left(X[l, m] * \Phi_k[l]\right) \bigg|_{l = 0},
\end{equation}
where $\Phi_k$ is the $k$'th MFFC triangular filter operating along the STFT index $l$. The evaluation at $l=0$ indicates that this operation is a multiplication and summation only.

We can view this process as a \ac{tf} decompistion with STFT filters $\upsilon_l \in \Upsilon$, which is modified with the Mel-scale triangular filters $\Phi_k$. $\upsilon_l$ provides time-shift invariance due to the STFT window, whereas $\Phi_k$ selectively averages frequency content of the STFT, thereby providing some frequency-shift invariance.

\section{Wavelets as a MFCC Generalisation}

We can construct a wavelet filterbank which, similar to MFCCs, introduces time and frequency-shift invariance in a similar manner to the Mel spectrogram. This process is constructed in the opposite order (frequency-shift, then time-shift invariance) to the Mel spectrogram. However, its properties remain similar \cite{ws}.

We construct wavelet band-pass filters $\upsilon_k \in \Upsilon$ which is spaced in frequency such that the desired invariance properties is obtained (as discussed in section \ref*{sec:p1i:wavelets}). We then utilise a single low-pass filter $\Phi[\vect{m}]$ which then provides invariance in the time and/or spatial dimensions:
\begin{gather}
    \label{eqn:general_scalogram}
    X[\vect{k}, \vect{m}] = \left|x * \upsilon_{\vect{k}}\right|_{\downarrow \vect{d}}, \ \upsilon_{\vect{k}} \in \Upsilon; \\
    \label{eqn:general_scattering}
    \tilde{X}[\vect{k}, \vect{m}] = X[\vect{k}, \vect{m}] * \Phi[\vect{m}]
\end{gather}

The notation in equations (\ref*{eqn:general_scalogram}) and (\ref*{eqn:general_scattering}) conforms the general \ac{tf} description of chapter \ref*{chap:p1i}, although the specific wavelet scattering operators in chapters \ref*{chap:p2} to \ref*{chap:p3} follow a different standard. These operators are known as wavelet scattering, which formalise the notation of ``averaging a \ac{tf}-decomposition'' in terms of digital signal processing. Chapters \ref*{chap:p2} to \ref*{chap:p3} discuss the specifics, flavours and implementation of scattering operators at length.

### 2024-05-14 dissertation/content/p3_intro.tex
started intro to p3

\chapter{Wavelet Scattering in Higher Dimensions}

>>>>

\chapter{Wavelet Scattering in Higher Dimensions}
\label{chap:p3i}

In chapter \ref*{chap:p2i}, we reviewed a generalised definition of a multi-dimensional TF decomposition $X$, which is processed by a set of invariance-enforcing filters $\Phi_i$. Wavelet scattering in 1 dimension utilising a single invariance-enforcing filter, $\phi$, as discussed in chapter \ref*{chap:p2}.

Wavelet scattering filterbanks have been extended to multiple dimensions using various different wavelets and filterbank construction techniques. This chapter provides an overview of the existing filterbanks and filters and their properties. We then identify a gap in the literature, which gives rise to a generalisation of cascasded 1D scattering transforms -- separable wavelet scattering.

\section{Types of Scattering Transforms}
Existing scattering transforms are listed below convenience:
\begin{enumerate}
    \item 1D scattering \citep{1dscattering1, ws}. This is the first-introduced scattering transform, which serves as a generalisation of MFCCs. It improves on MFCCs by recovering the information lost due to invariance filter (averaging) operations. Morlet wavelets are used, with cascaded scattering levels having its own number of filter per octave.
    \item 2D scattering \citep{2dscattering}. Gabor filters modified to 0-mean, are used to construct the filterbank. The filterbank consists of exponentially placed filters in one axis, which is then rotated by equally spaced angles to cover the frequency plane. This filterbank only has a single filter per octave ($Q=1$). In particular, in polar coordinates, filters are placed exponentially along $r$, but linearly along $\theta$.
    \item 3D scattering -- harmonic scattering \citep{harmonicscattering}. The spherical harmonics are used as filters which consitute an orthogonal basis in 3 dimensions. These filters have ``lobes'' that extend in a multitude of dimensions, depending on the harmonic, in the frequency domain.
    \item 3D scattering -- extension of 2D scattering \citep{3dscattering}. The filterbank is constructed similarly to 2D scattering, where Gabor filters are spaced with linearly varying angles (in spherical coordinates) in 3 dimensions.
    \item Joint time-frequency \citep{ws_joint_tf,jointtfscattering2} -- extending 1D to 2D. A first level of scattering in performed in 1D to produce a scalogram ($X[k, m]$). 2D scattering is then performed on subsequent levels, directly on the scalogram. The 2D wavelets are separable - a product of 1D wavelets. This allows 1D scattering on audio to have features that are also dependent on adjascent frequency bins. It is conceptually similar to a \ac{cnn} operating on a \ac{stft} spectrogram.
\end{enumerate}

\section{A Detailed view of Joint Time-Frequency Scattering}

\section{The Benefits of Generalising Joint Time-Frequency}
